---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First, we'll load the data.

I'm going to use the ReadR package to unzip the file, so I'll load that package.  I'll also load the tidyverse package at the same time.
```{r, message=FALSE}
require(readr)
require(tidyverse)
require(xtable)
```

```{r}
rm(list = ls())
```


Next, I"ll load the data.

```{r}
df <- read_csv("activity.zip")
head(df)
str(df)
```
The assignment specified that NA's could be ignored, but I wanted to look at how many there are.


## What is mean total number of steps taken per day?

```{r}
daily_total_steps <- df %>% 
  group_by(date) %>% 
  summarize(total = sum(steps, na.rm = TRUE)) 
daily_total_steps
```

```{r}
hist(daily_total_steps$total, breaks = 10, main = "Histogram of Daily Total Steps", xlab = "Steps")
rug(daily_total_steps$total)
```

```{r}
summary(daily_total_steps)
```

## What is the average daily activity pattern?

```{r}
intervals_df <- df %>% 
  group_by(interval) %>% 
  summarise(meansteps = mean(steps, na.rm = TRUE))

with(intervals_df, plot(x = interval, y = meansteps, type = "l",
                        xlab = "Interval during Day", 
                        ylab = "Average Steps in Interval",
                        main = "Average Daily Activity Pattern"))

```

```{r}
intervals_df %>% filter(meansteps == max(meansteps))
```

## Imputing missing values


```{r}
#creates a column to check for NA's in any of the columns
df.na <- df %>% 
  mutate(na_col = steps + date + interval)

sum(is.na(df.na$na_col))

#sums NAs in the steps column
sum(is.na(df$steps))
```

**We'll use the mean value for the interval across all days to fill in missing values.**  We've already calcualted this in the "intervals_df" dataframe.

```{r}
df.na <- merge(x = df, y = intervals_df, by = "interval", ) 

df.na[is.na(df.na$steps), "steps"] <- df.na[is.na(df.na$steps), "meansteps"]

head(df.na)

```

```{r}
daily_na_total_steps <- df.na %>% 
  group_by(date) %>% 
  summarize(total = sum(steps, na.rm = TRUE)) 
daily_total_steps

par(mfrow = c(1,2), mar = c(5, 4, 2, 1))
hist(daily_total_steps$total, breaks = 10, main = "Daily Total Steps", xlab = "Steps", ylim = c(0, 25))
rug(daily_total_steps$total)
hist(daily_na_total_steps$total, breaks = 10, main = "Daily Total Steps, Imputed NAs", xlab = "Steps", ylim = c(0, 25))
rug(daily_total_steps$total)
```

```{r}
summary(daily_total_steps$total)
summary(daily_na_total_steps$total)

#Change in Median
(10766-10395) / 10395
#Change in Mean
(10766-9354) / 9354
```

Imputing the NA values has a relatively small impact on the median (increases it by 3.6%), but a substantial effect on median, increasing it by 15%.

## Are there differences in activity patterns between weekdays and weekends?


```{r}
weekend_days <- c("Saturday", "Sunday")

df.na$weekend <- factor((weekdays(df.na$date) %in% weekend_days), levels = c(FALSE, TRUE), labels = c("weekday", "weekend"))

table(df.na$weekend)
```

```{r}
library(lattice)
```

```{r}
dayofweek_df <- df.na %>% 
  group_by(interval, weekend) %>% 
  summarise(total = mean(steps))
with(dayofweek_df,  xyplot(total ~ interval | weekend, layout = c(1, 2), type = "l"))
```

