---
title: "Assessment1_report.rmd"
author: "Zac"
date: "1/3/2021"
output: html_document
---

# Reproducible Research: Peer Assessment #1


### Loading and preprocessing the data.
First, we'll load the data.

I'm going to use the ReadR package to unzip the file, so I'll load that package.  I'll also load the tidyverse package at the same time.
```{r, message=FALSE}
require(readr)
require(tidyverse)
require(xtable)
```

```{r}
rm(list = ls())
```


Next, I"ll load the data.

```{r}
df <- read_csv("activity.zip")
head(df)
str(df)
```
The assignment specified that NA's could be ignored, but I wanted to look at how many there are.
```{r}
mean(is.na(df$steps))
```

### What is the mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

```{r}
daily_total_steps <- df %>% 
  group_by(date) %>% 
  summarize(total = sum(steps, na.rm = TRUE)) 
daily_total_steps
```
2. Create a histogram of total steps taken per day.

```{r}
hist(daily_total_steps$total, breaks = 10, main = "Histogram of Daily Total Steps", xlab = "Steps")
rug(daily_total_steps$total)
```


3. Calculate and report the mean and median of the total number of steps taken per day.

```{r}
summary(daily_total_steps)
```


### What is the average daily activity pattern?

1. Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days (y-axis)


```{r}
intervals_df <- df %>% 
  group_by(interval) %>% 
  summarise(meansteps = mean(steps, na.rm = TRUE))

with(intervals_df, plot(x = interval, y = meansteps, type = "l",
                        xlab = "Interval during Day", 
                        ylab = "Average Steps in Interval",
                        main = "Average Daily Activity Pattern"))

```



2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
intervals_df %>% filter(meansteps == max(meansteps))
```

### Imputing Missing Values

**Note that there are a number of days / intervals where there are missing values**  The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset.
```{r}
#creates a column to check for NA's in any of the columns
df.na <- df %>% 
  mutate(na_col = steps + date + interval)

sum(is.na(df.na$na_col))

#sums NAs in the steps column
sum(is.na(df$steps))
```

```{r}
head(df.na)

```

2. Devise a strategy for filling in all of the missing values in the dataset.  

**We'll use the mean value for the interval across all days to fill in missing values.**  We've already calcualted this in the "intervals_df" dataframe.

3. Create a new dataset that is equal to the original dataset, but with the missing data filled in.
```{r}
df.na <- merge(x = df, y = intervals_df, by = "interval", ) 

df.na[is.na(df.na$steps), "steps"] <- df.na[is.na(df.na$steps), "meansteps"]

head(df.na)

```
4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day.  Do these values differ from the estimates from the first part of the assignment?  What is the impact of imputing missing data on the estimates of the total daily number of steps?
  
```{r}
daily_na_total_steps <- df.na %>% 
  group_by(date) %>% 
  summarize(total = sum(steps, na.rm = TRUE)) 
daily_total_steps

par(mfrow = c(1,2), mar = c(5, 4, 2, 1))
hist(daily_total_steps$total, breaks = 10, main = "Daily Total Steps", xlab = "Steps", ylim = c(0, 25))
rug(daily_total_steps$total)
hist(daily_na_total_steps$total, breaks = 10, main = "Daily Total Steps, Imputed NAs", xlab = "Steps", ylim = c(0, 25))
rug(daily_total_steps$total)
```
```{r}
summary(daily_total_steps$total)
summary(daily_na_total_steps$total)

#Change in Median
(10766-10395) / 10395
#Change in Mean
(10766-9354) / 9354
```

Imputing the NA values has a relatively small impact on the median (increases it by 3.6%), but a substantial effect on median, increasing it by 15%.

### Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicated whether a given weekday or weekend day. (using the filled-in values datset).

```{r}
weekend_days <- c("Saturday", "Sunday")

df.na$weekend <- factor((weekdays(df.na$date) %in% weekend_days), levels = c(FALSE, TRUE), labels = c("weekday", "weekend"))

table(df.na$weekend)
```

2. Make a panel plot containing a time series plot (type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays and weekend days.  

```{r}
library(lattice)
```

```{r}
dayofweek_df <- df.na %>% 
  group_by(interval, weekend) %>% 
  summarise(total = mean(steps))
with(dayofweek_df,  xyplot(total ~ interval | weekend, layout = c(1, 2), type = "l"))
```

